Mamba-Based Tumor Segmentation\\ \thanks{$\dagger$ Equal contributions}

### 2024-04-26 paper/main.tex
finalized

Making use of Structured State Spaces, presented by Gu, Goel, and Ré \cite{gu2022efficiently}, for their superiority in long range dependency tasks, a specific type of SSM, the Mamba architecture, by Gu and Dao \cite{gu2023mamba}, proposed a foundational model to operate on arbitrary sequences from a variety of inputs in the domain of sequence modeling. This model achieves the power of Transformers while scaling linearly in sequencing length to be computationally efficient. Additionally, the selection mechanism of Mamba improves on an SSMs ability to focus on or ignore sections in an input sequence by parameterizing the parameters of an SSM to reflect the current context of the input. The Mamba model and its application to brain tumor MRI imaging segmentation will be evaluated in this project.

@@

For this our analysis, we compare four models against our own developed ones. The models are U-Net\cite{ronneberger2015unet}, ResNet18 \cite{he2015deep}, ResNet50, U-Mamba\cite{ma2024umamba}, and SegViT\cite{dosovitskiy2021image}. Below we explain our choices.
\subsection{U-Net}
The core idea behind U-Net is to complement a traditional contracting network with successive layers that replace pooling operations with upsampling operators. These upsampling layers aim to increase the resolution of the output. Subsequent convolutional layers can then learn to assemble a precise output based on this high-resolution information \cite{ronneberger2015unet}.
A key modification in the U-Net architecture is the inclusion of a large number of feature channels in the upsampling part. This allows the network to effectively propagate contextual information to higher resolution layers. As a result, the expansive path of the network becomes more or less symmetric to the contracting part, leading to a distinctive U-shaped architecture. To predict pixels at the border regions of the image, the missing context is extrapolated by mirroring the input image. This tiling strategy is crucial for applying the network to large images, as it circumvents resolution limitations imposed by GPU memory constraints \cite{ronneberger2015unet}. Due to its encoder-decoder architecture, skip connections, multi-scale feature extraction, and its efficiency, U-Net has been used in many denoising and diffusion models. For instance, DDPMs (Denoising Diffusion Probabilistic Models) use an architecture similar to U-Net for denoising and sample generation \cite{ho2020denoising} while the popular Stable Diffusion architecture uses a U-Net based architecture for the diffusion process \cite{rombach2021highresolution}. We have chosen to compare our novel segmentation architecture against the U-Net architecture due to its well-established reputation and widespread adoption in the field of medical image segmentation.

\subsection{ResNet}
We have also selected ResNet18 and ResNet50 as additional benchmarks. These architectures, introduced by He et al. in their seminal work "Deep Residual Learning for Image Recognition" \cite{he2015deep}, have revolutionized the field of deep learning by addressing the problem of vanishing gradients in deep neural networks. The key innovation in ResNets is the introduction of residual connections, which allow the network to learn residual functions with reference to the input layer, thereby facilitating the training of much deeper networks. ResNet18 and ResNet50, with 18 and 50 layers respectively, have been widely adopted in various computer vision tasks, including image classification, object detection, and segmentation. These models have demonstrated exceptional performance and generalization ability across diverse datasets. By comparing our proposed architecture against ResNet18 and ResNet50, we aim to assess its effectiveness in relation to these well-established and highly influential architectures. This comparison will provide valuable insights into the capabilities of our model and its potential to advance the state-of-the-art in image segmentation tasks.
\subsection{U-Mamba}
As our model takes much inspiration from the recently developed U-Mamba design \cite{ma2024umamba}, we also choose to incorporate it in our comparison. U-Mamba addresses these limitations by introducing a novel hybrid CNN-SSM block that leverages the strengths of both architectures. The convolutional layers in the block are responsible for local feature extraction, while the State Space Sequence Models (SSMs) \cite{gu2022efficiently}, a new family of deep sequence models, are known for their strong capability in handling long sequences and capturing long-range dependencies. By integrating these two components, U-Mamba achieves a balance between local and global information processing, enabling it to effectively handle long-range dependencies in biomedical image segmentation tasks. Moreover, U-Mamba incorporates a self-configuring mechanism that allows it to automatically adapt to various datasets without manual intervention, enhancing its versatility and usability.
\subsection{SegFormer}
SegFormer \cite{xie2021segformer} is a transformer-based semantic segmentation model that employs a hierarchical structure and a novel attention mechanism called Efficient Self-Attention (ESA). The SegFormer architecture consists of a transformer encoder for capturing long-range dependencies and a lightweight All-MLP decoder for generating high-resolution segmentation masks. The ESA module reduces the computational complexity of self-attention by performing attention operations in a local window and aggregating global information through a depth-wise convolution. This allows SegFormer to efficiently process high-resolution images while capturing both local and global context. Moreover, SegFormer introduces a position-sensitive embedding scheme that encodes both spatial and channel-wise information, enhancing the model's ability to capture fine-grained details. By comparing U-Mamba against SegFormer, we aim to evaluate the effectiveness of our hybrid CNN-SSM approach in relation to the transformer-based segmentation mechanism employed by SegFormer. This comparison will provide insights into the strengths and weaknesses of both architectures and their ability to handle complex biomedical image segmentation tasks. Furthermore, it will help us understand the potential of transformer-based approaches and hybrid approaches in advancing the state-of-the-art in biomedical image segmentation.

@@

\subsection{UL-VM-UNet-v2}
In UL-VM-UNet-v2, the number of parallel branches in the Parallel Vision Mamba (PVM) block was increased from 4 to 8. This modification directly decreases the parameter count, further proving the idea proposed in the UltraLight VM-UNet paper \cite{wu2024ultralight} that processing features in parallel with reduced channel counts can significantly reduce parameters while maintaining performance.

@@

\subsection{UL-VM-UNet-v4}
This model incorporates pyramidal pooling, similar to UMambaBot-PP, but within every PVM block in both the encoder and decoder in the UltraLight VM-UNet architecture. The goal was to leverage the benefits of multi-scale feature extraction while maintaining the parameter efficiency of the UltraLight VM-UNet.

\subsection{UL-VM-UNet-v5}
UL-VM-UNet-v5 is meant to be a comparison to the -v4 and -v6 UltraLight variations. This model trained and performed well with 1.1M parameters. Incorporating pyramidal pooling on the UL network scored slightly worse than without pyramidal pooling but required approximately 400,000 more parameters.

\subsection{UL-VM-UNet-v6}
Another variation to further test the effects of incorporating pyramidal pooling in the UltraLight VM-UNet architecture was made by only using pyramidal pooling in the bottleneck layer, similarly done in UMambaBot-PP. The main goal here, again, was to integrate pyramidal pooling for the benefit of multi-scale feature extraction while keeping a low parameter count.

@@

Our experiments and comparisons with state-of-the-art (SOTA) models demonstrate the potential of the Mamba-based approach in both model size and segmentation accuracy. The smallest model, ULVM-Net-v2, achieved a compelling dice score of 0.8243 with only 0.042M parameters, achieving comparable performance to the larger ResNet50 model (42.9M params) with a dice score of 0.8672, while also outperforming the original UltraLight VM-UNet in terms of both parameter count and dice score.

@@

\appendix
% Redefine the section numbering for the appendix (optional)
\renewcommand{\thesection}{\Alph{section}}
% First appendix section
\subsection{Training Pipeline}
The training pipeline for our model consists of the following steps:

1. Data preprocessing: The dataset is split into training, validation, and test sets. The images are resized to a uniform size and normalized.

2. Model architecture: We employ a convolutional neural network (CNN) with the following layers: [Describe the layers and their configurations].

3. Training: The model is trained using the training set for 100 epochs with a batch size of 32. We use the Adam optimizer with a learning rate of 0.0003.

4. Evaluation: The trained model is evaluated on the validation set to monitor its performance and prevent overfitting. We use the dice loss metric.

5. Testing: Once the model achieves satisfactory performance on the validation set, it is tested on the held-out test set to assess its generalization ability.

>>>>

Making use of Structured State Spaces, presented by Gu, Goel, and Ré \cite{gu2022efficiently}, for their capability in long range dependency tasks, a specific type of SSM, the Mamba architecture, by Gu and Dao \cite{gu2023mamba}, proposed a foundational model to operate on arbitrary sequences from a variety of inputs in the domain of sequence modeling. This model achieves the power of Transformers while scaling linearly in sequencing length to be computationally efficient. Additionally, the selection mechanism of Mamba improves on an SSMs ability to focus on or ignore sections in an input sequence by parameterizing the parameters of an SSM to reflect the current context of the input. The Mamba model and its application to brain tumor MRI imaging segmentation will be evaluated in this project.

@@

For this our analysis, we compare four models against our own developed ones. The models are U-Net\cite{ronneberger2015unet}, ResNet18 \cite{he2015deep}, ResNet50, U-Mamba\cite{ma2024umamba}, and SegViT\cite{dosovitskiy2021image}. Below we explain our choices.
\subsection{U-Net}
The core idea behind U-Net is to complement a traditional contracting network with successive layers that replace pooling operations with upsampling operators. These upsampling layers aim to increase the resolution of the output. Subsequent convolutional layers can then learn to assemble a precise output based on this high-resolution information \cite{ronneberger2015unet}.
A key modification in the U-Net architecture is the inclusion of a large number of feature channels in the upsampling part. This allows the network to effectively propagate contextual information to higher resolution layers. As a result, the expansive path of the network becomes more or less symmetric to the contracting part, leading to a distinctive U-shaped architecture. To predict pixels at the border regions of the image, the missing context is extrapolated by mirroring the input image. This tiling strategy is crucial for applying the network to large images, as it circumvents resolution limitations imposed by GPU memory constraints \cite{ronneberger2015unet}. Due to its encoder-decoder architecture, skip connections, multi-scale feature extraction, and its efficiency, U-Net has been used in many denoising and diffusion models. For instance, DDPMs (Denoising Diffusion Probabilistic Models) use an architecture similar to U-Net for denoising and sample generation \cite{ho2020denoising} while the popular Stable Diffusion architecture uses a U-Net based architecture for the diffusion process \cite{rombach2022highresolution}. We have chosen to compare our novel segmentation architecture against the U-Net architecture due to its well-established reputation and widespread adoption in the field of medical image segmentation.

\subsection{ResNet}
We have also selected ResNet18 and ResNet50 as additional benchmarks. These architectures, introduced by He et al. in their seminal work "Deep Residual Learning for Image Recognition" \cite{he2015deep}, have revolutionized the field of deep learning by addressing the problem of vanishing gradients in deep neural networks. The key innovation in ResNets is the introduction of residual connections, which allow the network to learn residual functions with reference to the input layer, thereby facilitating the training of much deeper networks. ResNet18 and ResNet50, with 18 and 50 layers respectively, have been widely adopted in various computer vision tasks, including image classification, object detection, and segmentation. These models have demonstrated exceptional performance and generalization ability across diverse datasets. By comparing our proposed architecture against ResNet18 and ResNet50, we aim to assess its effectiveness in relation to these well-established and highly influential architectures. This comparison will provide valuable insights into the capabilities of our model and its potential to advance the state-of-the-art in image segmentation tasks.
\subsection{U-Mamba}
As our models take inspiration from the recently developed U-Mamba design \cite{ma2024umamba}, we also choose to incorporate it in our comparison. U-Mamba addresses these limitations by introducing a novel hybrid CNN-SSM block that leverages the strengths of both architectures. The convolutional layers in the block are responsible for local feature extraction, while the State Space Sequence Models (SSMs) \cite{gu2022efficiently}, a new family of deep sequence models, are known for their strong capability in handling long sequences and capturing long-range dependencies. By integrating these two components, U-Mamba achieves a balance between local and global information processing, enabling it to effectively handle long-range dependencies in biomedical image segmentation tasks. Moreover, U-Mamba incorporates a self-configuring mechanism that allows it to automatically adapt to various datasets without manual intervention, enhancing its versatility and usability.
\subsection{SegFormer}
SegFormer \cite{xie2021segformer} is a transformer-based semantic segmentation model that employs a hierarchical structure and a novel attention mechanism called Efficient Self-Attention (ESA). The SegFormer architecture consists of a transformer encoder for capturing long-range dependencies and a lightweight All-MLP decoder for generating high-resolution segmentation masks. The ESA module reduces the computational complexity of self-attention by performing attention operations in a local window and aggregating global information through a depth-wise convolution. This allows SegFormer to efficiently process high-resolution images while capturing both local and global context. Moreover, SegFormer introduces a position-sensitive embedding scheme that encodes both spatial and channel-wise information, enhancing the model's ability to capture fine-grained details. By comparing U-Mamba against SegFormer, we aim to evaluate the effectiveness of our hybrid CNN-SSM approach in relation to the transformer-based segmentation mechanism employed by SegFormer. This comparison will provide insights into the strengths and weaknesses of both architectures and their ability to handle complex biomedical image segmentation tasks. Furthermore, it will help us understand the potential of transformer-based approaches and hybrid approaches in advancing the state-of-the-art in biomedical image segmentation.

@@

\subsection{UL-VM-UNet-v2}
In UL-VM-UNet-v2, the number of parallel branches in the Parallel Vision Mamba (PVM) block was increased from 4 to 8. This modification directly decreases the parameter count, further proving the idea proposed in the UltraLight VM-UNet paper \cite{wu2024ultralight} that processing features in parallel with reduced channel counts can significantly reduce parameters while maintaining performance. The channel list is the same as Ultra LightM-UNet's, [8,16,24,32,48,64].

@@

\subsection{UL-VM-UNet-v4}
This model incorporates pyramidal pooling, similar to UMambaBot-PP, but within every PVM block in both the encoder and decoder in the UltraLight VM-UNet architecture. The goal was to leverage the benefits of multi-scale feature extraction while maintaining the parameter efficiency of the UltraLight VM-UNet. Incorporating pyramidal pooling on the UL network scored slightly worse than without pyramidal pooling but required approximately 400,000 more parameters. The channel list was [16,32,64,128,256,512].

\subsection{UL-VM-UNet-v5}
UL-VM-UNet-v5 is meant to be a comparison to the -v4 and -v6 UltraLight variations. This model trained and performed well with 0.7M parameters. v5 has 8 parallels and modifies the channel list to [16,32,64,128,256,512].

\subsection{UL-VM-UNet-v6}
Another variation to further test the effects of incorporating pyramidal pooling in the UltraLight VM-UNet architecture was made by only using pyramidal pooling in the bottleneck layer, similarly done in UMambaBot-PP. The main goal here, again, was to integrate pyramidal pooling for the benefit of multi-scale feature extraction while keeping a low parameter count. The channel list was also [16,32,128,256,512].

@@

Our experiments and comparisons with SOTA models demonstrate the potential of the Mamba-based approach in both model size and segmentation accuracy. The smallest model, ULVM-Net-v2, achieved a compelling dice score of 0.8243 with only 0.042M parameters, achieving comparable performance to the larger ResNet50 model (42.9M params) with a dice score of 0.8672, while also outperforming the original UltraLight VM-UNet in terms of both parameter count and dice score.

@@

\appendix
% Redefine the section numbering for the appendix (optional)
\renewcommand{\thesection}{\Alph{section}}

% First appendix section
\subsection{Training}
We train multiple models for image segmentation using an 80-10-10 train-test-evaluation split. The accuracy of the models is evaluated using Dice scores, and the DiceLoss is used as the loss function.  \\
PP-UMamba, UMambaBot, ResNet18, ResNet50, and UNet were trained for 100 epochs using DiceLoss as the loss function and Adam as the optimizer with a learning rate of 3e-4.\\
UltraLight-VM-UNet and UltraLight-VM-UNet\_v1-v6 were trained for 100 epochs using BCEDiceLoss as the loss function and AdamW as the optimizer with a learning rate of 3e-4. All UltraLight variants (v1-v6) are trained using the same settings as UltraLight-VM-UNet. \\
SegFormer was trained for 100 epochs using BCEWithLogitsLoss as the loss function and AdamW as the optimizer with a learning rate of 1e-4. The learning rate is reduced on a plateau using a scheduler, and early stopping is employed. \\
More detailed training information can be found in our GitHub repository: \url{https://github.com/lxaw/mamba-tumor-seg/} and under the \texttt{tests} folder.
